---
title: "Introduction to msDiaLogue - Spring 2026"
subtitle: "UConn Statistical Consulting Services"
date: "2026-03-18"
author: "Timothy E. Moore, Eric Bae, & Shiying Xiao"
echo: true
format:
  revealjs:
    code-line-numbers: true
    slide-number: true
    css: styles.css
number-sections: false  # optional
crossref:
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

## 

**NOTE:** This file is available in the Day 3 folder of this workshop's OneDrive

```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(kableExtra)
library(gtsummary)

```

## Why msDiaLogue

To develop a flexible, customizable analysis pipeline for PMF users for use with proteomics experiments:

-   Primarily DIA-focussed, but expanded to DDA (Scaffold)
-   Adding peptide-level functionality (e.g., Phospo-proteome, Ubuquitome)
-   Including metabolomics

**We want analysis choices to be as transparent as possible**

See: <https://uconn-scs.github.io/msDiaLogue/>.

## Overview

Aim to be able to conduct analysis from start to finish within an integrated platform:

-   Transformation
-   Normalization
-   Imputation
-   DA Analysis
-   Summarization & Visualization

## Goals

-   To showcase some of the methodological features included in **msDiaLogue**.
-   Illustrate an analysis workflow on a 'real' dataset.
-   Promote the usage of this software for any proteomics (and metabolomics) analyses done in conjunction with PMF
-   Ask: any features you'd like to see?

## Example data

Set of plasma samples that were either not enriched (non-depleted), or enriched by one of 3 different bead types from various vendors:

::: {#tbl-1 .quarto-table tbl-cap="Example data structure"}
```{r echo = FALSE}
library(knitr)
library(kableExtra)

knitr::kable(
  data.frame(
    Condition = c("ENR", "ENRPlus", "ABC"),
    N = c(3, 3, 3)
  ),
  align = "l"
) %>%
  kable_paper("hover") %>%
  kable_styling(font_size = 40)
```
:::

## R programming basics

```{r, eval = FALSE, echo = TRUE}

object_1 <- function(argument1  = X)
  
object_2 <- function(argument_1 = "AAA", 
                   argument_2 = 123, 
                   argument_3  = object1)


```

## installation

```{r, eval = FALSE}

install.packages("pak")
pak::pkg_install("uconn-scs/msDiaLogue")

```

## preprocessing()

```{r, echo = FALSE}
dat_0 <- read.csv("data/20250921_182428_20250921_PAcer_Ceres_HumanPlasmaEnrichmentComparison_1hrDIAPASEF_Report.csv")

dat_0 <- dat_0[!dat_0$PG.Genes=="DMKN",]
dat_0 <- dplyr::filter(dat_0,!grepl("CON_", PG.ProteinAccessions))

dat_0 %>% group_by(R.Replicate, R.Condition) %>%slice_head(n=100) %>%
  write.csv(., "cleaned.csv")
  
  
```

```{r}
dataRaw <- preprocessing(dataSet = dat_0,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)

```

## preprocessing()

#### Original

```{r, echo = FALSE}

head(dat_0, 3)

```

#### Preprocessed

```{r, echo = FALSE}

head(datRaw[1:3,1:4])

```


## transform()

```{r}

trans <- transform(datRaw, method = "log", logFold = 2)

```


## normalize()

```{r}

norm <- normalize(trans, applyto = "sample", normalizeType = "median")

```

## filtering (again)

```{r}
filt <- filterNA(norm, minProp = 0.2, by = "cond", saveRm = TRUE)




```

## imputation

```{r}
visualize.dist(norm)

```


## imputation

```{r, results='hide'}

imp_full <- impute.knn_seq(norm)

imp_filt <- impute.knn_seq(filt)


```

## imputation

```{r}
visualize.dist(list("Pre-Imputation" = norm, 
                    "Post-Imputation" = imp_full))


```

## imputation

```{r}

visualize.dist(list("Pre-Imputation" = filt, 
                    "Post-Imputation" = imp_filt))


```

## summarization

```{r}
dataSumm <- summarize(imp_filt, saveSumm = TRUE)  %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))


dataSumm[1:5, 1:5]

```

## analysis

```{r}

levels(imp_filt$R.Condition)

anlys_modt <- analyze.mod_t(imp_filt, ref = "Nondepleted")

head(anlys_modt$`ENR-Nondepleted`[,1:5])


```

## analysis

### unsupervised

```{r}
anlys_pca <- analyze.pca(imp_filt, center = TRUE, scale = TRUE)

```

## analysis

### supervised & regularized
```{r}
anlys_plsda <- analyze.plsda(imp_filt, method = "kernelpls",
                             center = TRUE, scale = FALSE)

```


## visualization

```{r}

visualize.rank(imp_filt %>% dplyr::select(R.Condition, R.Replicate,
                                          KAIN_HUMAN, 
                                          sample(4:ncol(imp_filt), 50, replace = F)), 
               listName = "KAIN_HUMAN", color = "darkred")

```


## visualization

```{r}
visualize.test(anlys_modt$`ENR-Nondepleted`)

```

## visualization

```{r}
visualize.test(anlys_modt)

```


```{r}

visualize.volcano(anlys_modt$`ENR-Nondepleted`, P.thres = 0.05, F.thres = 1)

```

```{r}
visualize.score(anlys_pca, ellipse = TRUE,
                ellipse.level = 0.95, 
                label = TRUE)
```


```{r}
visualize.vip(anlys_plsda, comp = 1, num = 10,
              thres = 0.2)
```